---
title: "Would more tuning budget help?"
subtitle: "Analysis of the xgboost and rpf tuning archives on Bikeshare"
date: today
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

```{r setup}
library(data.table)
library(ggplot2)

theme_set(theme_minimal(base_size = 12) + theme(legend.position = "bottom"))

read_archive <- function(model) {
  archive <- fread(file.path("results", sprintf("%s_archive.csv", model)))
  archive[, model := model]
  archive[, incumbent := cummin(regr.mse)]
  archive[]
}

archives <- list(xgb = read_archive("xgb"), rpf = read_archive("rpf"))
both <- rbindlist(lapply(archives, \(a) a[, .(model, eval, init_design, regr.mse, incumbent, runtime_learners)]))
fold_cols <- c("mse_fold1", "mse_fold2", "mse_fold3")

# Bounds on the tuning scale: log(value) for logscale params
search_space <- fread("results/search_space.csv")
```

Both learners were tuned with asynchronous MBO, 1000 evaluations each, using
3-fold CV MSE.
The first ~100 evaluations are the random initial design; after that the
surrogate model proposes configurations.
All numbers here come from `results/*_archive.csv`, so this report renders
without the `_targets/` store.

## Incumbent trajectory

```{r fig-incumbent}
#| fig-cap: "MSE of evaluated configurations (points) and best-so-far MSE (line), zoomed to the range below the incumbent after the initial design. Dashed line: size of the initial design."
# Async evaluation interleaves late initial-design results with MBO proposals, so mark its size, not its last eval
n_init <- both[, .(n_init = sum(init_design)), by = model]
zoom <- both[n_init, on = "model"][, .(upper = incumbent[n_init[1]]), by = model]
zoomed <- both[zoom, on = "model"][, incumbent := pmin(incumbent, upper)]

ggplot(zoomed, aes(x = eval)) +
  geom_point(data = zoomed[regr.mse <= upper], aes(y = regr.mse), alpha = 0.3, size = 0.8) +
  geom_step(aes(y = incumbent), linewidth = 0.8, color = "#b2182b") +
  geom_vline(data = n_init, aes(xintercept = n_init), linetype = "dashed") +
  facet_wrap(vars(model), scales = "free_y") +
  labs(x = "Evaluation", y = "3-fold CV MSE")
```

```{r tab-budget}
budget_fractions <- c(0.1, 0.25, 0.5, 0.75, 1)
budget <- both[,
  .(
    fraction = budget_fractions,
    eval = ceiling(budget_fractions * .N),
    incumbent = incumbent[ceiling(budget_fractions * .N)]
  ),
  by = model
]
budget[, `:=`(
  rmse = sqrt(incumbent),
  gain_vs_final_pct = 100 * (incumbent / min(incumbent) - 1)
), by = model]

knitr::kable(
  budget,
  digits = c(0, 2, 0, 1, 2, 2),
  col.names = c("Model", "Budget fraction", "Evaluation", "Best MSE", "Best RMSE", "MSE above final best (%)")
)

second_half <- budget[fraction %in% c(0.5, 1), .(gain = 100 * (1 - min(incumbent) / max(incumbent))), by = model]
gain <- setNames(second_half$gain, second_half$model)
```

The second half of the budget lowered the best MSE by
`r sprintf("%.1f%%", gain[["xgb"]])` for xgboost and
`r sprintf("%.1f%%", gain[["rpf"]])` for rpf.
In RMSE (the unit of the target, bikers per hour) that is well below one
biker.

## Are the remaining gains distinguishable from CV noise?

The best configuration is the minimum of 1000 noisy CV estimates, so it is
optimistically biased: part of its lead over the runner-up is luck.
Two noise scales to compare against:

- **Standard error of the CV mean** of the best configuration (fold SD / √3).
  This is roughly how much its estimate would move under a fresh resampling,
  and it dwarfs the differences between top configurations.
- **Paired differences** to the best configuration.
  All configurations share the same CV folds (instantiated once per tuning
  run), so per-fold differences cancel the fold effect.
  With only 3 folds, the standard error of that difference rests on 2 degrees
  of freedom and is itself very noisy.

```{r tab-noise}
noise <- rbindlist(lapply(archives, \(a) {
  best <- a[which.min(regr.mse)]
  best_folds <- unlist(best[, ..fold_cols])
  top <- a[order(regr.mse)][2:21]
  paired_diff <- as.matrix(top[, ..fold_cols]) - matrix(best_folds, nrow(top), 3, byrow = TRUE)
  data.table(
    model = best$model,
    best_mse = best$regr.mse,
    fold_sd = sd(best_folds),
    cv_se = sd(best_folds) / sqrt(3),
    gap_to_20th = top[.N, regr.mse] - best$regr.mse,
    median_paired_se = median(apply(paired_diff, 1, sd) / sqrt(3)),
    n_within_1pct = a[regr.mse <= 1.01 * best$regr.mse, .N]
  )
}))

knitr::kable(
  noise,
  digits = c(0, 1, 1, 1, 1, 1, 0),
  col.names = c(
    "Model", "Best MSE", "Fold SD of best", "SE of best's CV mean", "Gap best to 20th best",
    "Median paired SE (top 20)", "Configs within 1% of best"
  )
)
```

```{r fig-folds}
#| fig-cap: "Per-fold MSE of the 10 best configurations. Lines connect the folds of one configuration; parallel lines mean the fold effect dominates the configuration effect."
top_folds <- rbindlist(lapply(archives, \(a) {
  top <- a[order(regr.mse)][1:10, c("model", "eval", "regr.mse", fold_cols), with = FALSE]
  top[, rank := seq_len(.N)]
  melt(top, measure.vars = fold_cols, variable.name = "fold", value.name = "mse")
}))
top_folds[, fold := sub("mse_", "", fold)]

ggplot(top_folds, aes(x = fold, y = mse, group = eval, color = rank)) +
  geom_line(alpha = 0.7) +
  geom_point() +
  facet_wrap(vars(model), scales = "free_y") +
  scale_color_viridis_c(direction = -1) +
  labs(x = NULL, y = "MSE", color = "Rank")
```

## Search space boundaries

If good configurations crowd against a bound, the optimum may lie outside the
search space, and more budget inside it will not find it.
Positions are normalized to [0, 1] within the bounds, on log scale where the
parameter was tuned on log scale.

```{r fig-bounds}
#| fig-cap: "Position of the 50 best configurations within each parameter's search range. 0 = lower bound, 1 = upper bound. Red: best configuration."
top_n <- 50
positions <- rbindlist(lapply(archives, \(a) {
  top <- a[order(regr.mse)][seq_len(top_n)]
  top[, rank := seq_len(.N)]
  space <- search_space[model == a$model[1]]
  long <- melt(top, id.vars = c("model", "rank"), measure.vars = space$param, variable.name = "param")
  long[, param := as.character(param)]
  long <- long[space, on = c("model", "param")]
  long[, position := (fifelse(logscale, log(value), value) - lower) / (upper - lower)]
}))

ggplot(positions, aes(x = position, y = param)) +
  geom_jitter(data = positions[rank > 1], height = 0.15, width = 0, alpha = 0.4) +
  geom_point(data = positions[rank == 1], color = "#b2182b", size = 3) +
  facet_wrap(vars(model), scales = "free_y", ncol = 1) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "Position in search range", y = NULL)
```

```{r tab-bounds}
near_bound_threshold <- 0.05
near_bound <- positions[,
  .(
    best_value = format(signif(value[rank == 1], 3)),
    share_near_lower = mean(position <= near_bound_threshold),
    share_near_upper = mean(position >= 1 - near_bound_threshold)
  ),
  by = .(model, param)
][share_near_lower >= 0.25 | share_near_upper >= 0.25]

knitr::kable(
  near_bound,
  digits = 2,
  col.names = c("Model", "Parameter", "Best value", "Share near lower", "Share near upper"),
  caption = sprintf(
    "Parameters where at least 25%% of the top %d configurations lie within %d%% of a bound.",
    top_n, 100 * near_bound_threshold
  )
)
```

Categorical parameters of rpf among the top configurations:

```{r tab-categorical}
top_rpf <- archives$rpf[order(regr.mse)][seq_len(top_n)]
knitr::kable(
  top_rpf[, .N, by = .(split_structure, delete_leaves)][order(-N)],
  caption = sprintf("Top %d rpf configurations", top_n)
)
```

## Cost

```{r tab-cost}
cost <- both[, .(
  total_cpu_hours = sum(runtime_learners) / 3600,
  median_eval_minutes = median(runtime_learners) / 60,
  best_eval_minutes = runtime_learners[which.min(regr.mse)] / 60
), by = model]

knitr::kable(
  cost,
  digits = 1,
  col.names = c("Model", "Total learner CPU hours", "Median minutes per eval", "Minutes for best config")
)
```

`runtime_learners` is the summed train and predict time over the 3 folds of one
evaluation.

## Conclusion

```{r conclusion-numbers}
second_half_mse <- both[, .(gain = incumbent[floor(.N / 2)] - min(incumbent)), by = model]
ratio <- setNames(noise$cv_se / second_half_mse$gain[match(noise$model, second_half_mse$model)], noise$model)
```

- **xgboost: converged.**
  The second half of the budget lowered the best MSE by
  `r sprintf("%.1f%%", gain[["xgb"]])`, about
  `r sprintf("%.0f", ratio[["xgb"]])`× smaller than the standard error of the
  best configuration's CV estimate.
  Hundreds of configurations are within 1% of the best.
  The parameters near a bound (`alpha`, `gamma`) sit at the lower end of
  log-scaled regularization ranges starting at 1e-3, which already means
  "practically no regularization"; widening those ranges would not change the
  model.
- **rpf: the search space, not the budget, is the limit.**
  Its second-half gain (`r sprintf("%.1f%%", gain[["rpf"]])`) is also about
  `r sprintf("%.0f", ratio[["rpf"]])`× below the CV standard error, but nearly
  all top configurations sit at the upper bound of `splits` and the lower bound
  of `t_try`.
  The optimum likely lies outside the current box.
  Figure 1 agrees: unlike xgboost, rpf's MBO proposals are still drifting
  downward at the end of the run, as the surrogate keeps pushing toward those
  bounds.
  A short follow-up tuning with a wider range for those two parameters (and
  `split_structure = "leaves"`, `delete_leaves = FALSE` fixed) is worth more
  than extending this run.
  Larger `splits` makes each evaluation slower, so expect higher cost per
  evaluation than the median above.
- **Either way:** the ranking among top configurations is mostly CV noise and
  selection bias.
  Re-evaluating the top 10–20 with repeated CV is a cheaper way to a more
  reliable pick than more search.
